Modelling fine-scale correlation structures using hidden Markov models

### 2020-07-13 Report/Sections/discussion.tex
Small Edits

We presented a collection of HMM models which can be combined together in increasingly complex hierarchical models to match the complexity of particular problems faced by practitioners. Traditional HMMs can be used to model a state-switching process with conditionally independent observations and Markovian dynamics within the hidden state. However, many real-world processes are more complicated than this.

For observations that do not appear to be conditionally independent in time, the CarHMM is a good choice. The CarHMM explicitly models auto-correlation in the emission distributions of the HMM while maintaining the structure needed to evaluate the likelihood using the forward algorithm. In our normal model formulation, we have added only one additional parameter, $\phi^{(i)}$, per possible hidden state \citep{Lawler:2019}. In addition, we prove in the appendix that under certain circumstances the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process. Several useful model selection tools such as the lag plot can test if there is significant auto-correlation within the observation sequence.

For simultaneous observed processes taking place at different time scales, the HHMM \citep{Barajas:2017,Adam:2019} utilizes hierarchical structures to jointly model both as HMMs. A coarse-scale HMM is assumed to emit both an observation $Y_t$ as well as another fine-scale HMM with hidden states $X^*_t$ and observations $Y^*_t$.

Finally, for processes with very high sampling frequencies and/or with intricate fine-scale structure, the HHMM structure developed by \citep{Barajas:2017,Adam:2019} can be generalized so the fine-scale model can be any model which admits an easy-to-calculate likelihood. For example, if the sampling rate of the fine-scale process is very high, then the fine-scale model can be a simple probability distribution over the summary statistics of a moving window of observations. In addition, researchers can recursively stack hierarchical HMM models together like building blocks to create increasingly complex models. This should be done with care, as it is important to balance the need to effectively capture the process in question with the need to avoiding over-fitting and slow parameter estimation. One way to temper model complexity is to reduce the dimension of the parameter space by forcing fine-scale states to be shared across the coarse-scale states, but model complexity inevitably grows rapidly as hierarchical structures are stack on top of each other.

An example of balancing model complexity with fast model fitting is presented in the simulation study here. While (fine-scale) sub-dive behavioral states were shared across (coarse-scale) dive types to reduce model complexity for the hierarchical models, by far the fastest model to train (~15 minutes) was the simple CarHMM-DFT. This models had no hierarchical component but near-perfect accuracy when decoding sub-dive behavioral states of the simulated whale. This model would be the preferred one if simultaneous modeling of dive types was not required. However, this model is deficient if ecologists wish to understand to joint relationship between dive type and intra-dive behaviour.

The simulation study also shows that the observed fisher information serves as a suitable approximation for the standard errors of parameter estimates in most cases. One notable exception is that the standard errors of the probability transition matrix estimates, $\hat \Gamma$ and $\hat \Gamma^*$, tend to be overestimated by the observed fisher information.

Finally, we used the CarHHMM-DFT to model the behavior of a killer whale off the coast of British Columbia, Canada. The CarHHMM-DFT was able to distinguish three distinct sub-dive behaviors and two dive types simultaneously due to its hierarchical structure. The DFT component proved useful in determining the sub-dive behaviors of the whale, as the mean of the emission distribution of $Z^{*(2)}$ was separated by an order of magnitude between each sub-dive state. Finally, the learned auto-correlation parameter for $\mathbf{Z}^{*(1)}$, $\phi$, was above 0.5 for every dimension and sub-dive type, providing evidence that the conditionally auto-regressive component of the CarHHMM-DFT resulted in a better fit of the data. The introduction of the parameter $\phi$ allows $\mathbf{Z}^{*(1)}$ to be interpreted as a state-switching OU process (see appendix).

>>>>

We presented a collection of HMM models which can be combined together in increasingly complex hierarchical models to match the complexity of particular problems faced by practitioners. This flexible framework can be used to deal with complicated dependence structures within time-series data.

Traditional HMMs can be used to model a state-switching process with conditionally independent observations and Markovian dynamics within the hidden state. However, many real-world processes are more complicated than this and require more complex models.

The CarHMM generalizes the HMM by explicitly modeling auto-correlation in the emission distributions of the HMM while maintaining the structure needed to evaluate the likelihood using the forward algorithm. In our normal model formulation, we have added only one additional parameter, $\phi^{(i)}$, per possible hidden state \citep{Lawler:2019}. Several useful model selection tools such as the lag plot can test if there is significant auto-correlation within an observation sequence.

Although the CarHMM can incorporate auto-correlation into the structure of an HMM, it can break down when observations are taken at irregular time intervals. A common solution to this issue is to use a continuous-time method such as the state-switching OU processes described by \citep{Michelot:2019}. Most continuous time models require relatively slow MCMC algorithms to perform inference and as a result are not easily incorporated into the HHMM structure. However, we prove that certain continuous-time methods are equivalent to an CarHMM under certain conditions (see appendix).

For simultaneous observed processes taking place at different time scales, the HHMM as described by \citep{Barajas:2017,Adam:2019} utilizes hierarchical structures to jointly model both as HMMs. In particular, each hidden state of a coarse-scale HMM is assumed to emit both an observation $Y_t$ as well as another fine-scale HMM with hidden states $X^*_t$ and observations $Y^*_t$.

Finally, for processes with very high sampling frequencies and/or with intricate fine-scale structure, the HHMM structure developed by \citep{Barajas:2017,Adam:2019} can be generalized so the fine-scale model can be any model which admits an easy-to-calculate likelihood. For example, if the sampling rate of the fine-scale process is very high, then the fine-scale model can be described by a simple probability distribution over the summary statistics of a moving window of observations. In addition, researchers can recursively stack hierarchical HMM models together like building blocks to create increasingly complex models. This should be done with care, as it is important to balance the need to effectively capture the process in question with the need to avoiding over-fitting and slow parameter estimation.

One way to temper model complexity is to reduce the dimension of the parameter space by forcing fine-scale states to be shared across the coarse-scale states. Even still, model complexity inevitably grows rapidly as hierarchical structures are stack on top of each other.

An example of balancing model complexity with fast model fitting is presented in the simulation study. While (fine-scale) sub-dive behavioral states were shared across (coarse-scale) dive types to reduce model complexity for the hierarchical models, by far the fastest model to train (\~ 15 minutes) was the CarHMM-DFT. The CarHMM-DFT had no hierarchical component but near-perfect accuracy when decoding sub-dive behavioral states of the simulated whale. This model would be preferred if simultaneous modeling of dive types was not required. However, this model is not sufficient if ecologists wish to understand to joint relationship between dive type and intra-dive behaviour.

The simulation study also shows that the observed fisher information serves as a suitable approximation for the standard errors of parameter estimates in most cases. One notable exception is that the standard errors of the probability transition matrix estimates ($\hat \Gamma$ and $\hat \Gamma^*$) tend to be overestimated by the observed fisher information.

Finally, we used the CarHHMM-DFT to model the behavior of a killer whale off the coast of British Columbia, Canada. The CarHHMM-DFT was able to distinguish three distinct sub-dive behaviors and two dive types simultaneously due to its hierarchical structure. The DFT component proved useful in determining the sub-dive behaviors of the whale, as the mean of the emission distribution of $Z^{*(2)}$ was separated by an order of magnitude between each sub-dive state. Finally, the learned auto-correlation parameter for $\mathbf{Z}^{*(1)}$, $\phi^*$, was above 0.5 for every dimension and sub-dive type, providing evidence that the conditionally auto-regressive component of the CarHHMM-DFT resulted in a better fit to the data. The introduction of the parameter $\phi^*$ also allows $\mathbf{Z}^{*(1)}$ to be interpreted as a state-switching OU process (see appendix).

### 2020-07-13 Report/Sections/models.tex
Small Edits

%\section{Models and parameter estimation}
%\input Sections/model_Nancy.tex

In this section we first remind the reader of the definition of a hidden Markov model (HMM). We then review three variations on the HMM; the CarHMM, HHMM, and HMM-DFT. Finally, we generalize the hierarchical structure of these models and describe how they can be flexibly altered and combined to form new models.

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=5in]{../Plots/HMM.png}
	\caption{Graphical representation of a traditional HMM.}
	\label{fig:HMM}
\end{figure}

@@

A key assumptions of an HMM is \textit{conditional independence} between observations given the state sequence.  Therefore, traditional HMMs do not hold when the observations exhibit certain forms of significant correlation in time.

The CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by \citep{Lawler:2019}, explicit models auto-correlation within an HMM. Like a traditional HMM, A CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$. Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$, conditional on $X_1,\ldots, X_T$ and $ \{Y_1,\ldots, Y_{t-1}\}$, depends on \textit{both} $X_t$ \textit{and} $Y_{t-1}$. 
The first emission $Y_1$ is assumed to be fixed as an initial value which does not depend upon $X_1$. (Fig \ref{fig:CarHMM}) shows the structure of a CarHMM.

\begin{figure}[ht]
	\centering
	\includegraphics[width=5in]{../Plots/CarHMM.png}
	\caption{Graphical representation of a traditional CarHMM.}
	\label{fig:CarHMM}
\end{figure}

%
We denote the conditional distribution of $Y_t$ given $Y_{t-1}= y_{t-1}$ and $ X_t=i$ as $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}( \cdot | y_{t-1})$.
As an example, we could assume that this conditional distribution is normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
%
\[
\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) = \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}
\]
and
\[
\mathbb{V}(Y_t| Y_{t-1} =y_{t-1}, X_t = i) = (\sigma^{(i)})^2.
\]
%
The likelihood for the CarHMM can be easily calculated using the forward algorithm. As previously, let $y$ be the vector of observed emissions. Then
\begin{equation}
\calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\Theta) \mathbf{1}_N
\label{CarHMM_likelihood}
\end{equation}
where
%
$P(y_t|y_{t-1};\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
%
\subsubsection{Equivalence with continuous-time process}

Although the CarHMM can incorporate auto-correlation into the structure of an HMM, they break down when observations are taken at irregular time intervals since the probability transition matrix $\Gamma$ would then have to depend upon the time step between observations. A common solution to this issue is to use a continuous-time method such as the one introduced by Michelot et al \citep{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying behavioural state of the animal. Namely, a state-switching Ornstein-Uhlenbeck process $Y$ is the solution to the following stochastic differential equation:
%
\begin{equation}
    \label{eqn:SDE}
    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t
\end{equation}
%
where $X_t$ follows some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process. One notable difference between the state-switching OU process and the CarHMM is that the time index $t \in \mathbb{R}$ exists in continuous time and is not necessarily an integer. If the behavioral state $X_t$ is known and does not change between observations, the solution to (eqn \ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
%
\begin{align}
Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
\label{eqn:OU_sol}
\end{align}
%
where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$. Continuous time models can be built up from arbitrarily complex stochastic differential equations and allow for uneven step lengths in the observations sequence $Y$. However, most continuous time models require MCMC algorithms to perform inference and as a result are not easily incorporated into the HHMM structure.

Although most continuous time models are difficult to incorporate into an HHMM, we show in that under certain conditions, the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process shown in (eqn \ref{eqn:OU_sol}). This gives new interpretation to the learned parameters of the CarHMM in the context of a continuous-time model.

\begin{theorem}{Theorem 1.}{}%
If:
\begin{enumerate}
    \item The hidden behavioural process $X$ from (Eq. \ref{eqn:SDE}) follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$, and:
    \item Observations of the SDE from (Eq. \ref{eqn:SDE}) are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
\end{enumerate}
then the observations $y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:

@@

A \textit{hierarchical hidden Markov model} contains both a coarse-scale process and a fine-scale process, both of which are modeled by an HMM. The coarse-scale process is a hidden Markov model as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain and $Y_1,\ldots, Y_T$ are the corresponding observed responses. The possible states are $1,\ldots, N$.   
%
In the hierarchical setting, each state $X_t$ emits not only $Y_t$ but also another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of fine-scale observed emissions, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$. The length of each fine-scale process, $T^*_t$, is a problem-specific tuning parameter. Conditional on the coarse-scale hidden state $X_t$, the joint fine-scale process, ($X_t^*, Y_t^*$), is a hidden Markov model with parameters depending on the value of $X_t$.  Specifically, given that $X_t=i$, $X_t^*$ is a Markov chain with states $1,\ldots, N_t^*$. The distribution of $X_t^*$, conditional on $X_t=i$, is given by an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial probability, denoted by the $N_t^*$-vector $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain. For simplicity, we take $N_t^* \equiv N^*$ although this is not necessary.
%
The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{(i,i^*)}$, and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or sometimes simply denoted by $f^{*(i,i^*)}(\cdot)$. Let $\Theta^{*(i)}=\left(\theta^{(i,1)}, \ldots, \theta^{(i,N^*)}\right)$, the fine-scale emission parameter vector corresponding to $X_t=i$.
%
In terms of the dependence structure, given the coarse-scale states, $X_1,\ldots, X_T$, the $T$ fine-scale processes $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are independent HMMs. Depending upon the process being modeled, it is possible to force certain parameters to be shared across different coarse or fine states. For example, in the killer whale case study in section \ref{sec:case_study} we force the fine-scale emission parameters to be shared across coarse-scale hidden states, i.e. $\theta^{(1,i^*)} = \ldots = \theta^{(N,i^*)}$ for $i^* = 1, \ldots, N^*$. (Fig \ref{fig:HHMM}) shows the dependence structure for an HHMM.

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=5in]{../Plots/HHMM.png}
	\caption{Graphical representation of a traditional HHMM.}
	\label{fig:HHMM}
\end{figure}

@@

The HMM with discrete Fourier transform or HMM-DFT, is a variation of the hierarchical hidden Markov model where the fine-scale process is no longer modeled with an HMM and instead summarized using it Fourier transform. For simplicity of presentation, we assume that the length of the fine-scale processes is constant, i.e. that $T^*_t = T^*$, although this need not be the case in general. Suppose that the fine-scale process $y^*_t$ does not switch states hidden states, but it does exhibit significant periodic behavior which cannot be modeled with a Markov chain. Then, we suggest using the discrete Fourier transform (DFT) on $y^*_t$:
%
\begin{align*}
    DFT\{y^*_t\}(k) := \hat{y}^{*(k)}_{t} = \sum_{t^* = 1}^{T^*} y^*_{t,t^*}\exp\left(-i \frac{2\pi k}{T^*} (t^*-1)\right) \qquad k = 0, 1, \ldots, T^*-1
\end{align*}
%
Summary statistics can then drastically reduce the dimension of $\hat{y}^*_t$. One example of summary statistics include the following:
%
\begin{equation}
    \label{eqn:z}
    z_t^{*(1)} := \mathcal{R}\left(\hat{y}^{(0)}_t\right) \qquad z_t^{*(2)} := \frac{1}{T^*}\sum_{k=1}^{\tilde{\omega}}|\hat{y}^{(k)}_t|^2
\end{equation}
%
$z_t^{*(1)}$ is equal to the average value of $y^*_t$ and $z_t^{*(2)}$ is equal to the squared 2-norm of the component of $y^*_t$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window length $T^*$. The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with care. These summary statistics are just one possible choice to describe each window; other choices include the dominant frequency and amplitude of $y^*_t$. Figure (\ref{fig:fourier_example}) visually displays the process of transforming $y^*$ into $z^*$.

\begin{figure}[t]
	\centering
	\includegraphics[width=5in]{../Plots/fourier_transform.png}
	\caption{Visualization of transforming $\mathbf{y^*}_t$ into $\mathbf{z}_t$ using a discrete Fourier transform.}
	\label{fig:fourier_example}
\end{figure}

Once $z^*_t$ is calculated, it can simply be treated as an observation of the coarse-scale HMM and incorporated into the emission distribution $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or simply $f^{(i)}\left(y_t,z^*_t\right)$. In total, the likelihood of the hierarchical HMM-DFT is as follows:
\begin{equation}
\calL_{\text{HMM-DFT}}(y,z^*;\Theta,\Gamma) = \delta P(y_1,z^*_1;\Theta) \prod_{t=2}^T \Gamma P(y_t,z^*_t;\Theta) \mathbf{1}_N
\label{HMMDFT_likelihood}
\end{equation}
where $P(y_t,z^*_t;\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$.

@@

In addition to the DFT and HMM models described above, the fine-scale process $Y^*_t$ can be modeled with \textbf{any} parametric model which admits an easy-to-compute likelihood. The state-specific emission distribution $\calL_{\text{HMM}}$ from the HHMM likelihood is then replaced by the likelihood of the fine-scale model, $\calL_{\text{fine}}(\mathbf{y}^*_t;\Theta^{*(i)})$:
\[
\calL_{\text{coarse}}(y,y^*;\Theta,\Theta^*,\Gamma) = \delta P(y_1,y^*_1;\Theta,\Theta^*) \prod_{t=2}^T \Gamma P(y_t,y^*_t;\Theta,\Theta^*) \mathbf{1}_N
\]
where $P(y_t,y^*_t;\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$. This definition is straightforward to extend to the CarHMM as well:
\[
\calL_{\text{coarse}}(y,y^*;\Theta,\Theta^*,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) \mathbf{1}_N
\]
where $P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$

As previously mentioned, the fine-scale model can be any parametric model with easy-to-compute likelihood, including any of the four models described in the previous four subsections (HMM, CarHMM, HHMM, and HMM-DFT). These HMM models then act as building blocks that can be used to build increasingly complex hierarchical models based on the HMMs. In the sections that follow, we perform both a simulation study and real-world case study modelling killer whale dive behaviour using models built from these building blocks.

>>>>

%\section{Models and parameter estimation}

In this section we first remind the reader of the definition of a hidden Markov model (HMM). We then review three variations on the HMM; the CarHMM, HHMM, and HMM-DFT. We also briefly describe a continuous-time model and its connection with the CarHMM. Finally, we generalize the hierarchical structure of these models and describe how they can be flexibly altered and combined to form new models.

@@

@@

A key assumption of an HMM is \textit{conditional independence} between observations given the state sequence.  Therefore, traditional HMMs do not hold when the observations exhibit certain forms of significant correlation in time.

The CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by \citep{Lawler:2019}, explicitly models auto-correlation within an HMM. Like a traditional HMM, A CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$. Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$, conditional on $X_1,\ldots, X_T$ and $ \{Y_1,\ldots, Y_{t-1}\}$, depends on \textit{both} $X_t$ \textit{and} $Y_{t-1}$. 
The first emission $Y_1$ is assumed to be fixed as an initial value which does not depend upon $X_1$. (Fig \ref{fig:CarHMM}) shows the structure of a CarHMM.

%
We denote the conditional distribution of $Y_t$ given $Y_{t-1}= y_{t-1}$ and $ X_t=i$ as $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}( \cdot | y_{t-1})$.
As an example, we could assume that this conditional distribution is normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
%
\[
\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) = \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}
\]
and
\[
\mathbb{V}(Y_t| Y_{t-1} =y_{t-1}, X_t = i) = (\sigma^{(i)})^2.
\]
%
The likelihood for the CarHMM can be easily calculated using the forward algorithm. As previously, let $y$ be the vector of observed emissions. Then
\begin{equation}
    \calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\Theta) \mathbf{1}_N
    \label{CarHMM_likelihood}
\end{equation}
where
%
$P(y_t|y_{t-1};\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
%
\subsection{Continuous-time processes}

While CarHMMs can model auto-correlation within the observation sequence, they fail when observations are taken at irregular time intervals. Another way to model a stochastic process is to use a stochastic differential equation, or SDE. As an example, \citep{Michelot:2019} model the movement of an animal as the solution to the following stochastic differential equation:
%
\begin{equation}
    \label{eqn:SDE}
    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t
\end{equation}
%
where $X_t$ follows some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process. This SDE is referred to as a state-switching Ornstein-Uhlenbeck process. Unlike HMMs, the time index $t \in \mathbb{R}$ exists in continuous time and is not necessarily an integer. If the behavioral state $X_t$ is known and does not change between observations, the solution to (eqn \ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
%
\begin{align}
    Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
    \label{eqn:OU_sol}
\end{align}
%
where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$. Most continuous time models are difficult to incorporate into an HHMM and require MCMC methods to fit. However, under certain conditions, the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process. This gives new interpretation to the learned parameters of the CarHMM in the context of a continuous-time model.

\begin{theorem}{Theorem 1.}{}%
If:
\begin{enumerate}
    \item The hidden behavioural process $X$ from (eq. \ref{eqn:SDE}) follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$, and:
    \item Observations of the SDE from (eq. \ref{eqn:SDE}) are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
\end{enumerate}
then the observations $Y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:

@@

A \textit{hierarchical hidden Markov model} contains both a coarse-scale process and a fine-scale process, both of which are HMMs. The coarse-scale process is a hidden Markov model as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observed responses.   
%
In the hierarchical setting, each state $X_t$ emits another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of fine-scale observed emissions, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$. This fine-scale process can be either instead of or in addition to the coarse-scale observations $Y_t$. The length of each fine-scale process is denoted as $T^*_t$ and can often be controlled by the practitioner. Conditional on the coarse-scale hidden state $X_t$, the joint fine-scale process, ($X_t^*, Y_t^*$), is a hidden Markov model with parameters depending on the value of $X_t$. Specifically, given that $X_t=i$, $X_t^*$ is a Markov chain with states $1,\ldots, N_t^*$. The distribution of $X_t^*$, conditional on $X_t=i$, is given by an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial probability, denoted by the $N_t^*$-vector $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain. For simplicity, we take $N_t^* \equiv N^*$ although this is not necessary.
%
The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{(i,i^*)}$, and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$. Let $\Theta^{*(i)}=\left(\theta^{(i,1)}, \ldots, \theta^{(i,N^*)}\right)$, the fine-scale emission parameter vector corresponding to $X_t=i$.
%
Given the coarse-scale states, $X_1,\ldots, X_T$, the $T$ fine-scale processes $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are independent HMMs. Depending upon the process being modeled, it is possible to force certain parameters to be shared across different coarse or fine states. For example, in the killer whale case study in section \ref{sec:case_study} we force the fine-scale emission parameters to be shared across coarse-scale hidden states, i.e. $\theta^{(1,i^*)} = \ldots = \theta^{(N,i^*)}$ for $i^* = 1, \ldots, N^*$. (Fig \ref{fig:HHMM}) shows a graphical representation of the dependence structure for an HHMM.

@@

@@

The HMM with discrete Fourier transform, or HMM-DFT, incorporates hierarchical structure into an HMM differently than an HHMM. In particular, the fine-scale process is no longer modeled with an HMM and instead summarized using its Fourier transform. For simplicity of presentation, we assume that the length of the fine-scale processes is constant, i.e. that $T^*_t = T^*$, although this need not be the case in general. Suppose that the fine-scale process $y^*_t$ does not switch states hidden states, but does exhibit significant periodic behavior. Then, we suggest using the discrete Fourier transform (DFT) on $y^*_t$:
%
\begin{align*}
    DFT\{y^*_t\}(k) := \hat{y}^{*(k)}_{t} = \sum_{t^* = 1}^{T^*} y^*_{t,t^*}\exp\left(-i \frac{2\pi k}{T^*} (t^*-1)\right) \qquad k = 0, 1, \ldots, T^*-1
\end{align*}
%
Summary statistics can then drastically reduce the dimension of $\hat{y}^*_t$. One such example is as follows:
%
\begin{equation}
    \label{eqn:z}
    z_t^{*(1)} := \mathcal{R}\left(\hat{y}^{(0)}_t\right) \qquad z_t^{*(2)} := \frac{1}{T^*}\sum_{k=1}^{\tilde{\omega}}|\hat{y}^{(k)}_t|^2
\end{equation}
%
$z_t^{*(1)}$ is the average value of $y^*_t$ and $z_t^{*(2)}$ is the squared 2-norm of the component of $y^*_t$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window length $T^*$. The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with care. These summary statistics are just one possible choice to describe each window; other choices include the dominant frequency and amplitude of $y^*_t$. Figure (\ref{fig:fourier_example}) visually displays the process of transforming $y^*$ into $z^*$.

Once $z^*_t$ is calculated, it can simply be treated as an observation of the coarse-scale HMM and incorporated into the emission distribution $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or more succinctly $f^{(i)}\left(y_t,z^*_t\right)$. In total, the likelihood of the hierarchical HMM-DFT is as follows:
\begin{equation}
    \calL_{\text{HMM-DFT}}(y,z^*;\Theta,\Gamma) = \delta P(y_1,z^*_1;\Theta) \prod_{t=2}^T \Gamma P(y_t,z^*_t;\Theta) \mathbf{1}_N
    \label{HMMDFT_likelihood}
\end{equation}
where $P(y_t,z^*_t;\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$.

@@

In addition to the DFT and HMM models described above, the fine-scale process $Y^*_t$ can be modeled with \textit{any} parametric model which admits an easy-to-compute likelihood. The state-specific emission distribution $\calL_{\text{HMM}}$ from the HHMM likelihood is then replaced by the likelihood of the fine-scale model, $\calL_{\text{fine}}(\mathbf{y}^*_t;\Theta^{*(i)})$:
\[
\calL_{\text{coarse}}(y,y^*;\Theta,\Theta^*,\Gamma) = \delta P(y_1,y^*_1;\Theta,\Theta^*) \prod_{t=2}^T \Gamma P(y_t,y^*_t;\Theta,\Theta^*) \mathbf{1}_N
\]
where $P(y_t,y^*_t;\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$. This definition is straightforward to extend to the CarHMM as well:
\[
\calL_{\text{coarse}}(y,y^*;\Theta,\Theta^*,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) \mathbf{1}_N
\]
where $P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$.

Possible candidates for the fine-scale model include any of the models described in the previous subsections (HMM, CarHMM, state-switching OU, HHMM, and HMM-DFT). These HMM models can act as building blocks which can be used to construct increasingly complex hierarchical models based on the HMMs. In the sections that follow, we perform both a simulation study and real-world case study modelling killer whale dive behaviour using models constructed from these building blocks.
